---
title: "Analysis of MK2 transcript levels in TCGA-LUAD (NSCLC) dataset"
author: "Karthik Suresh"
date: "May 15, 2019"
output:
  html_document:
    theme: united
    toc: yes
  pdf_document:
    toc: yes
---
Here, we are downloading clinical data from TCGA-LUAD dataset using the TCGAbiolinks API. For the MK2 transcript data, I grabbed these data from oncoLnc.org using "MAPKAPK2" as the gene name and TCGA-LUAD as the cohort. 

# Data retrieval and cleaning

## Retrieve data from TCGA dataset, _join it with the MK2 transcript data

We retrieve the clinical data directly using GDC_query. For the MK2 transcript data, this was downloaded from OncoLnc.org using the following parameters: Gene-MAPKAPK2, and percentile: 50/50. We did this because this allowed us to get transcript levels on all available patients (since upper and lower 50% would include all patients). We then calculated the quantiles manually, so that in the future, if we needed to change the threshold, we could do so without re-downloading the OncoLnc data. 

```{r setup, include=TRUE}
knitr::opts_chunk$set(echo = TRUE)
rm(list=ls())
library("SummarizedExperiment")
library("dplyr")
library("DT")
library("TCGAbiolinks")
library(survival)
library(ggplot2)
library(ggpubr)
library(survminer)
library(knitr)
library("tidyr")
library("ggpmisc")

clinical <- GDCquery_clinic(project = "TCGA-LUAD", type = "clinical")
expr <- read.csv("../rawdata/LUAD.MAPKAPK2Exp.csv")

# the left_join. we need the clinical data, so we left_joined by TCGA patient id to the clinical dataset. This means that if there was a patient with clinical data without transcript levels, that would be a 'NA'. 
clinical <- left_join(clinical, expr, c("submitter_id"= "Patient"))

# let's do some checking to make sure the left_join happened properly, since this is a critical step. 

# So, here, what we have done is to collect the pt. id and MK2 expression data from 2 datasets: the joined clinical data frame, which contains the Expr data and the clinical data, and the MK2 Expression data set, which was downloaded from OncoLnc.org and only contains the downloaded Expr data. Next for each clinical pt, we asked if the Expression value in the joined data matches the Expression value from the other dataset. 

JoinCheck_clin <<- cbind(clinical$submitter_id, clinical$Expression)
JoinCheck_oncolnc <<- cbind(as.character(expr$Patient), expr$Expression)

colnames(JoinCheck_clin) <- c("clin.id", "clin.val")

colnames(JoinCheck_oncolnc) <- c("oncolnc.id", "oncolnc.val")

JoinCheck_clin<-as.data.frame(JoinCheck_clin)
JoinCheck_oncolnc<-as.data.frame(JoinCheck_oncolnc)

CheckValue <- function(clin.id, clin.val) {
   if(!is.na(clin.val)){
      
      if(JoinCheck_oncolnc[JoinCheck_oncolnc$oncolnc.id==clin.id,]$oncolnc.val==clin.val) return(1)
      else return(0)
   }
}

JoinCheck_clin$checkVal <- mapply(CheckValue, as.character(JoinCheck_clin$clin.id), JoinCheck_clin$clin.val)

# the checkVal value should be either 1 (the joined Expression value and the downloaded Expression value match) or NULL (clinical data is present, but there is no expression value found in the oncolnc data).
table(as.character(JoinCheck_clin$checkVal))


```

## Clean the data, create a df for an LR and Cox PH analyses later

Now that we have left-joined the data, and checked to make sure that the join was done properly, we move on to prep the data for Cox PH work. 
```{r include=TRUE}


clinical$time <- mapply(function(x,y) if(is.na(x)) return(y) else return(x), clinical$days_to_death, clinical$days_to_last_follow_up) 
clinical$age_at_diagnosis <- clinical$age_at_diagnosis/365
clinical$race <- as.factor(clinical$race)
clinical$smoking <- lapply(clinical$cigarettes_per_day, function(x) if(!is.na(x)) return(1) else return(0))
clinical$tumor_stage <- as.factor(clinical$tumor_stage)
clinical$vital_status <- as.factor(clinical$vital_status)
clinical$status <- lapply(clinical$vital_status, function(x) if(x=="alive") return(0) else return(1))
clinical$smoking <- as.numeric(clinical$smoking)


# combine clinical stage

combineStage <- function(x) {
if(!is.na(x))  {
if(x=="Stage I" | x=="Stage IB" | x=="Stage IA") return("Early Stage")
if(x=="Stage II" | x=="Stage IIA" | x=="Stage IIB") return("Early Stage")
if(x=="Stage IIIA" | x=="Stage IIIB") return("Late Stage")
if(x=="Stage IV") return("Late Stage")
}
  } 

# combined histologic types 

combinePath <- function(x) {
 if(!is.na(x))  {
   
   if(x=="Adenocarcinoma, NOS") return("Adenocarcinoma")
   if(x=="Mucinous adenocarcinoma")  return("Adenocarcinoma")
   if(x=="Adenocarcinoma with mixed subtypes") return("Adenocarcinoma")
   if(x=="Papillary adenocarcinoma") return("Adenocarcinoma")
   else return("Non-Adenocarcinoma")
   
 }
  
}

# combine sub-stages into 1,2,3
clinical$ajcc_pathologic_stage_combined <- lapply(clinical$ajcc_pathologic_stage, combineStage)

clinical$ajcc_pathologic_stage_combined <- as.factor(as.character(clinical$ajcc_pathologic_stage_combined))

# combine histology types
clinical$primary_diagnosis_combined <- lapply(clinical$primary_diagnosis, combinePath)
clinical$primary_diagnosis_combined<- as.factor(unlist(clinical$primary_diagnosis_combined))

# get rid of NULL entries
clinical <- clinical[clinical$ajcc_pathologic_stage_combined!="NULL",]

# we create a restricted dataset that only contains the variables we're interested in for CoxPH
MK2_CoxPH <-data.frame(clinical$age_at_diagnosis, clinical$gender, clinical$smoking, clinical$submitter_id, clinical$ajcc_pathologic_stage_combined, clinical$year_of_diagnosis, clinical$primary_diagnosis_combined, clinical$Expression, clinical$time, clinical$vital_status, clinical$age_at_diagnosis)

# We already got rid of NULLs for ajcc_pathologic_stage. However, the factor levels still list 3 levels - early, late and NULL. To re-level, convert to character, re-convert to factor, and now only 2 levels. 
MK2_CoxPH$clinical.ajcc_pathologic_stage_combined<-as.factor(as.character(MK2_CoxPH$clinical.ajcc_pathologic_stage_combined))

MK2_CoxPH <- MK2_CoxPH[!is.na(MK2_CoxPH$clinical.Expression), ]

# create censor variable for CoxPH
MK2_CoxPH$censor <- lapply(MK2_CoxPH$clinical.vital_status, function(x) if(x=="Alive") return(0) else return(1))
MK2_CoxPH$censor <- as.numeric(MK2_CoxPH$censor)
```


# Exploratory graphs

## Plot differences in transcript level by tumor histology and cancer stage

Basic plots looking at MK2 expression in 'Early' vs 'Late' in the LUAD dataset. 
```{r}
# plot MK2 differences for adenocarcinoma only
ggplot(clinical[clinical$primary_diagnosis_combined=="Adenocarcinoma", ], aes(x=ajcc_pathologic_stage_combined, y=Expression, ))+geom_boxplot()+geom_point(position=position_jitter())+stat_compare_means()+theme_bw()+ylab("MK2 transcript level")


ggplot(clinical, aes(x=ajcc_pathologic_stage_combined, y=Expression, fill=primary_diagnosis_combined))+geom_boxplot()+geom_point(position=position_jitterdodge())+theme_bw()

# Anova and Tukey HSD
NSCLC_MK2_aov<- aov(data=clinical, Expression~primary_diagnosis_combined + ajcc_pathologic_stage_combined+ajcc_pathologic_stage_combined*primary_diagnosis_combined)

TukeyHSD(NSCLC_MK2_aov)
```


# Cox PH model modeling Survival ~ MK2 expression + covariates

## Continuous MK2 expression:  Survival ~ expression (continuous) + histologic stage (Adeno vs. non-adeno) + clinical stage (early v. late)

In this model, we use MK2 expression as a continuous variable - in other words, don't split into low vs. high (for that, see below). The 'clinical.Expression' variable is the MK2 transcript level. 
```{r}


# COX PH model
MK2_coxph_model <- coxph(Surv(clinical.time, censor) ~ clinical.Expression+clinical.ajcc_pathologic_stage_combined+clinical.primary_diagnosis_combined+clinical.age_at_diagnosis, id=clinical.submitter_id, data=MK2_CoxPH)

# examine MK2 expression values 
hist(MK2_CoxPH$clinical.Expression)


# PH assumption testing

# we cacn see that statistical testing of schoenfield residuals is non-significant (i.e. PH assumption not violated)
PHtesting_MK2contin <- cox.zph(MK2_coxph_model)
PHtesting_MK2contin

# residual examination for individual variables
ggcoxdiagnostics(MK2_coxph_model, type="dfbeta")
ggcoxzph(PHtesting_MK2contin, font.main=c(8,"plain", "black"), font.x=c(8, "plain", "black"), font.y=c(8, "plain", "black"), font.caption=c(8, "plain", "black"), font.tickslab=c(8, "plain", "black"))

summary(MK2_coxph_model)
exp(confint(MK2_coxph_model))
```

## COX PH looking at survival ~ expression when expression is categorized as upper 33% or lower 33%.

Mechanistically, we're more interested in whether high vs. low MK2 transcript levels impact outcome. For this, we first categorize MK2 transcript levels based on quantiles and only take the lower and upper third in our groups for comparison. 
```{r}

# classify the MK2 expression variable as either top or bottom quantile, based on quantile definitions

bottom_quant <<- 0.33
top_quant <<- 0.66

MK2_CoxPH$MK2_Expression_bottomQ <- lapply(MK2_CoxPH$clinical.Expression, function (x) if(x < quantile(MK2_CoxPH$clinical.Expression, bottom_quant))  return(1) else return(0))

MK2_CoxPH$MK2_Expression_topQ <- lapply(MK2_CoxPH$clinical.Expression, function (x) if(x > quantile(MK2_CoxPH$clinical.Expression, top_quant))  return(1) else return(0))

# this is now a new data frame that only includes patients whose MK2 levels fall in either the bottom or the top quantile

MK2_CoxPH_tb <- MK2_CoxPH[MK2_CoxPH$MK2_Expression_bottomQ ==1 | MK2_CoxPH$MK2_Expression_topQ==1,]
MK2_CoxPH_tb$MK2_Expression_topQ <- as.numeric(MK2_CoxPH_tb$MK2_Expression_topQ)
MK2_CoxPH_tb$MK2_Expression_bottomQ <- as.numeric(MK2_CoxPH_tb$MK2_Expression_bottomQ)

# re-categorize the variable as MK2_tv 
MK2_CoxPH_tb$MK2_tv <- mapply(function(bottomQ, topQ) if(bottomQ==1) return("Low") else return("High"), MK2_CoxPH_tb$MK2_Expression_bottomQ, MK2_CoxPH_tb$MK2_Expression_topQ)
MK2_CoxPH_tb$MK2_tv <- as.factor(MK2_CoxPH_tb$MK2_tv)

# Cox PH for univariate model using MK2 as a categorical variable
MK2_coxph_model_tb <- coxph(Surv(clinical.time, censor) ~ MK2_tv, id=clinical.submitter_id, data=MK2_CoxPH_tb)

PHtesting_MK2categ<- cox.zph(MK2_coxph_model_tb)
ggcoxzph(PHtesting_MK2categ)
# this is not as useful because topQ is a categorical variable. a log-log plot will be more useful
ggcoxdiagnostics(MK2_coxph_model_tb, type="dfbeta")
# log-log plot showing lack of proportionality
plot(survfit(Surv(MK2_CoxPH_tb$clinical.time, MK2_CoxPH_tb$censor)~as.numeric(MK2_CoxPH_tb$MK2_tv)), fun="cloglog", xlab="log(t)", ylab="log(-log(Survival))") 

# Cox PH for multivariate model using  MK2 as a categorical variable
MK2_coxph_model_tb_mv <- coxph(Surv(clinical.time, censor) ~ MK2_tv+clinical.gender+clinical.age_at_diagnosis+clinical.smoking, id=clinical.submitter_id, data=MK2_CoxPH_tb)

summary(MK2_coxph_model_tb_mv)

```

## Using a time variable to account for PH assumption violation

We have a categorical variable (MK2 quantile, 'High' vs. 'Low') that violates PH. We split the time into chunks. We will arbitrary cut for the step function at t=100,300, 600,1000d. 

```{r}

MK2_tvary <- survSplit(Surv(clinical.time, censor) ~ ., data=MK2_CoxPH_tb, cut=c(365, 730,1095), episode="tgroup", zero=0)

# have a look at the result of split by time chunks at 0-365d (tgroup=1), 365-720d (tgroup=2) and so on...
kable(MK2_tvary[1:20,])

# model with time and stage stratified
MK2_tvary_model <- coxph(Surv(tstart,clinical.time, censor)~MK2_tv:strata(tgroup):strata(clinical.ajcc_pathologic_stage_combined), data=MK2_tvary)
MK2_tvary_model
exp(confint(MK2_tvary_model))


MK2_tvary$clinical.smoking <- as.factor(as.numeric(MK2_tvary$clinical.smoking))

MK2_tvary_model2 <- coxph(Surv(tstart,clinical.time, censor)~MK2_tv:strata(tgroup):strata(clinical.ajcc_pathologic_stage_combined)+clinical.gender+clinical.age_at_diagnosis.1+clinical.smoking, data=MK2_tvary)
MK2_tvary_model2
exp(confint(MK2_tvary_model2))

# this suggests that the effect is limited to a time period between 100-300 days. 


```

# Supplementary Analyses

## Jackknife-type analysis for effects due to long period of patient recruitment

It appears that patients were recruited across many years. Chemo regimens change with time. To Figure out whether this plays a role, we re-estimate the Early Stage 1st time strata point estimate (0.27, p = 0.02 in the multivariate model above) multiple times, each time removing only the patients recruited in a particular year
```{r}

years <- as.data.frame(table(as.factor(as.character(clinical$year_of_diagnosis))))
MK2_tvary$clinical.year_of_diagnosis <- as.factor(as.character(MK2_tvary$clinical.year_of_diagnosis))


returnPointEst <- function(x) {
   
   modelresult <- coxph(Surv(tstart,clinical.time, censor)~MK2_tv:strata(tgroup):strata(clinical.ajcc_pathologic_stage_combined)+clinical.gender+clinical.age_at_diagnosis.1+clinical.smoking, data=MK2_tvary[MK2_tvary$clinical.year_of_diagnosis!=x,])
   return(exp(modelresult$coefficients[4]))
}

years$estimate <- lapply(as.character(years$Var1), returnPointEst)
years$estimate <- lapply(years$estimate, function(x) return(x[[1]]))

returnPointEstlci <- function(x) {
   
   modelresult <- coxph(Surv(tstart,clinical.time, censor)~MK2_tv:strata(tgroup):strata(clinical.ajcc_pathologic_stage_combined)+clinical.gender+clinical.age_at_diagnosis.1+clinical.smoking, data=MK2_tvary[MK2_tvary$clinical.year_of_diagnosis!=x,])
   return(exp(confint(modelresult)[[4]]))
}

years$estimate_lci <- lapply(as.character(years$Var1), returnPointEstlci) 

returnPointEstuci <- function(x) {
   
   modelresult <- coxph(Surv(tstart,clinical.time, censor)~MK2_tv:strata(tgroup):strata(clinical.ajcc_pathologic_stage_combined)+clinical.gender+clinical.age_at_diagnosis.1+clinical.smoking, data=MK2_tvary[MK2_tvary$clinical.year_of_diagnosis!=x,])
   return(exp(confint(modelresult)[[23]]))
}

years$estimate_uci <- lapply(as.character(years$Var1), returnPointEstuci) 

years$estimate <- as.numeric(years$estimate)
years$estimate_lci<- as.numeric(years$estimate_lci)
years$estimate_uci <- as.numeric(years$estimate_uci)

ggplot(years, aes(x=Var1, y=estimate))+geom_point()+geom_errorbar(aes(ymin=estimate_lci, ymax=estimate_uci), width=0)+theme_bw()+ylim(0,1.2)+geom_hline(yintercept = 1)+ylab("HR for hi MK2 at 0-365d in multivariate model")+xlab("Patient Recruitment Year that was excluded from model")+theme(axis.text.x = element_text(angle = 90, hjust = 1))

```


## Differences in follow-up time. 
There was a concern raised re: differences in follow-up time between the early/late group. Here we look at that. 
```{r}
ggplot(data=MK2_CoxPH_tb, aes(fill=MK2_tv, y=clinical.time, x=clinical.ajcc_pathologic_stage_combined, ))+geom_boxplot()+geom_point(position=position_jitterdodge() )+theme_bw()+facet_wrap(~clinical.vital_status)

ggplot(data=MK2_CoxPH_tb, aes(x=clinical.time,fill=MK2_tv))+geom_density(alpha=0.3)+theme_bw()
ggplot(data=MK2_CoxPH_tb[MK2_CoxPH_tb$clinical.ajcc_pathologic_stage_combined=="Early Stage",], aes(x=clinical.time, fill=MK2_tv))+geom_density(alpha=0.3)+theme_bw()
ggplot(data=MK2_CoxPH_tb[MK2_CoxPH_tb$clinical.ajcc_pathologic_stage_combined=="Late Stage",], aes(x=clinical.time, fill=MK2_tv))+geom_density(alpha=0.3)+theme_bw()


clinical$smoking <- as.factor(as.numeric(clinical$smoking))

```

## Relationship between caspase-3, Mk2 and YAP1 transcript levels. 

The question now is whether caspase-3/YAP transcript levels correlate with MK2 transcript levels. 

```{r}
# relationship between casp-3 and MK2 and YAP1

casp3Exp <- read.csv("../rawdata/LUAD.casp3Exp.csv")
YAPExp <- read.csv("../rawdata/LUAD.YapExp.csv")
Casp3AndMK2 <- left_join(expr, casp3Exp, by=c("Patient"))
YAPAndMK2 <- left_join(expr, YAPExp, by=c("Patient"))
Casp3AndMK2$Expression.MK2.scaled <- scale(Casp3AndMK2$Expression.x)
Casp3AndMK2$Expression.Casp3.scaled <- scale(Casp3AndMK2$Expression.y)


YAPAndMK2$Expression.MK2.scaled <- scale(YAPAndMK2$Expression.x)
YAPAndMK2$Expression.YAP.scaled <- scale(YAPAndMK2$Expression.y)

ggplot(Casp3AndMK2, aes(x=Expression.MK2.scaled, y=Expression.Casp3.scaled))+geom_point()+theme_bw()+geom_smooth(method="lm")

ggplot(YAPAndMK2, aes(x=Expression.MK2.scaled, y=Expression.YAP.scaled))+geom_point()+theme_bw()+geom_smooth(method="lm")

```

# Second round of analysis - after speaking with Kevin.

We reviewed the Cox PH Schoenfield diagnostics. It became clear that PH violation was largely due to patients who were censored out to several years after diagnosis. Since we're really just interested in the first year, we opted to simply just censor at year one. Then, we model either  death at 1 year using logistic regression or time to death (censored at 1 year) using Cox PH. 

Another idea that emerged was look at the proportion of people who died in one year by MK2 deciles or quartiles. We can then regress this relationship.  

With regards to the prolonged enrollment time, rather than jackknife, we chose to now perform a sensitivity analysis for  time of enrollment/time since diagnosis

The transcript level units is in 1000s - Kevin suggested we scale this down by a factor of 1000 so the point estimate units are in line with the units for the other measurements. 

## Changes in proportion of people who died across quantiles of MK2
```{r}
# quantile and logistic regression

# this is the new data frame for this new round of calculations. 

MK2_CoxPH_quantile <- MK2_CoxPH
# the scaling here is to help the point estimates for the LR and Cox PH models be more reasonable. 
MK2_CoxPH_quantile$clinical.Expression.scaled <- MK2_CoxPH_quantile$clinical.Expression/1000

# this provides the variable for R censoring at one year and the outcome variable for logit regression.
MK2_CoxPH_quantile$death_at_1year <- mapply(function(dead,time) if(dead==1 & time <366) return(1) else if(dead==0 & time > 366) return(0) else if (dead==0 & time < 366) return(0) else if (dead==1 & time > 366) return(0), MK2_CoxPH_quantile$censor, as.numeric(MK2_CoxPH_quantile$clinical.time))

# this piece of code above is critical - so let's double check and make sure everything got done right

ggplot(MK2_CoxPH_quantile, aes(x=clinical.submitter_id, y=clinical.time, color=as.factor(clinical.vital_status)))+ geom_point()+facet_wrap(~as.factor(death_at_1year))+ylim(0,500)+xlab("Patient ID") + ylab("time to death")+geom_hline(yintercept = 365)
```

Interpretation off the plot above. So, we have colored each patient’s vital status by alive or dead, and we have faceted by whether they were assigned as “Death in one year” or not. So, if our code worked, all patients who died in less than one year (blue color; y < 365) should be classified as dead at one year (“1”). All patients who were alive in that time period should be classified as red and in the “0” column. After 365 days, everyone should have gotten a classification of “0” for death at one year, since they are all by definition still alive at 1 year, regardless of subsequent alive or dead status (color should mix, all colors in the “0” column). And that’s what we see…

```{r}
#Now, moving on.

MK2_CoxPH_quantile$MK2_quantile <- ntile(MK2_CoxPH_quantile$clinical.Expression.scaled, 5)
a <- aggregate(death_at_1year~MK2_quantile+clinical.ajcc_pathologic_stage_combined, FUN=sum, data=MK2_CoxPH_quantile)
b <- aggregate(death_at_1year~MK2_quantile+clinical.ajcc_pathologic_stage_combined, FUN=length, data=MK2_CoxPH_quantile)

quantile_df<-  as.data.frame(a$MK2_quantile)
quantile_df$stage <- as.data.frame(a$clinical.ajcc_pathologic_stage_combined)
quantile_df$prop <- as.data.frame(unlist(a$death_at_1year/b$death_at_1year))
quantile_df$prop <- unlist(quantile_df$prop)
colnames(quantile_df) <- c("quantile", "stage", "prop")
p1 <- ggplot(quantile_df[quantile_df$stage=="Early Stage",], aes(x=quantile, y=prop))+geom_point()+geom_smooth(method="lm", se=FALSE)+stat_poly_eq(formula="y~x", aes(label = paste(..eq.label.., ..rr.label.., sep = "~~~")), parse = TRUE)+theme_bw()+ylim(0,0.5)

p2 <- ggplot(quantile_df[quantile_df$stage=="Late Stage",], aes(x=quantile, y=prop))+geom_point()+geom_smooth(method="lm", se=FALSE)+stat_poly_eq(formula="y~x", aes(label = paste(..eq.label.., ..rr.label.., sep = "~~~")), parse = TRUE)+theme_bw()+ylim(0,0.5)

# This graph shows the relationship between MK2 quantile  and proportion of patients  who died in Early and Late Stage cancers. The relationship between MK2 quantile and Proportion of those patients  who died (denominator: all patients  in that quantile) 
gridExtra::grid.arrange(p1,p2, ncol=2)

```

## Logistic regression modeling death at 1 year ~ MK2+covariates

```{r}
MK2_CoxPH_quantile$death_at_1year <- as.factor(MK2_CoxPH_quantile$death_at_1year)
# logistic regression and ROC
MK2_CoxPH_quantile$clinical.ajcc_pathologic_stage_early_stage <- unlist(MK2_CoxPH_quantile$clinical.ajcc_pathologic_stage_early_stage)
MK2_CoxPH_quantile$MK2_quantile <- as.factor(MK2_CoxPH_quantile$MK2_quantile)

MK2_logreg1 <- glm(death_at_1year ~ clinical.Expression.scaled, data=MK2_CoxPH_quantile[MK2_CoxPH_quantile$clinical.ajcc_pathologic_stage_combined], family=binomial(link="logit"))
MK2_logreg1 <- glm(death_at_1year ~ MK2_quantile, data=MK2_CoxPH_quantile[MK2_CoxPH_quantile$clinical.ajcc_pathologic_stage_combined=="Late Stage",], family=binomial(link="logit"))

summary(MK2_logreg1)






K2_logreg4 <- glm(death_at_1year~ MK2_quantile, data=MK2_CoxPH_quantile, family=binomial(link="logit"))
MK2_logreg5 <- glm(death_at_1year~ MK2_quantile+clinical.ajcc_pathologic_stage_combined, data=MK2_CoxPH_quantile, family=binomial(link="logit"))
MK2_logreg6 <- glm(death_at_1year~ MK2_quantile+clinical.ajcc_pathologic_stage_combined+MK2_quantile*clinical.ajcc_pathologic_stage_combined, data=MK2_CoxPH_quantile, family=binomial(link="logit"))

MK2_logreg7 <- glm(death_at_1year~ MK2_quantile+clinical.ajcc_pathologic_stage_combined, data=MK2_CoxPH_quantile, family=binomial(link="logit"))

MK2_logreg8 <- glm(death_at_1year ~ clinical.Expression.scaled+clinical.smoking+clinical.age_at_diagnosis.1+clinical.gender, data=MK2_CoxPH_quantile[MK2_CoxPH_quantile$clinical.ajcc_pathologic_stage_combined=="Early Stage",], family=binomial(link="logit"))
MK2_logreg9 <- glm(death_at_1year ~ clinical.Expression.scaled+clinical.ajcc_pathologic_stage_combined, data=MK2_CoxPH_quantile, family=binomial(link="logit"))

MK2_logreg9 <- glm(death_at_1year ~ clinical.Expression.scaled, data=MK2_CoxPH_quantile[MK2_CoxPH_quantile$clinical.ajcc_pathologic_stage_combined=="Late Stage",], family=binomial(link="logit"))

# sensitivity analysis - add time since enrollment


ResourceSelection::hoslem.test(MK2_logreg3$y, fitted(MK2_logreg3))
summary(MK2_logreg3)
exp(MK2_logreg3$coefficients)
exp(confint.lm(MK2_logreg3))
```
## Logistic model with Caspase-3 added in as a variable

```{r}
colnames(casp3Exp) <- c("casp3.Patient", "casp3.time_to_death", "casp3.vital.status", "casp3.Expression", "casp3.Group")
MK2_CoxPH_MK2andCasp3 <- left_join(MK2_CoxPH_quantile, casp3Exp, by=c("clinical.submitter_id"="casp3.Patient"))
MK2_CoxPH_MK2andCasp3$casp3.Expression.scaled <- MK2_CoxPH_MK2andCasp3$casp3.Expression/1000
MK2_CoxPH_MK2andCasp3$casp3.Expression.scaled <- as.numeric(as.character(MK2_CoxPH_MK2andCasp3$casp3.Expression.scaled))

MK2_CoxPH_MK2andCasp3$MK2_quantile <- as.factor(MK2_CoxPH_MK2andCasp3$MK2_quantile)
MK2_logreg10 <- glm(death_at_1year ~ MK2_quantile+clinical.ajcc_pathologic_stage_combined+casp3.Expression.scaled, data=MK2_CoxPH_MK2andCasp3, family=binomial(link="logit") )

ResourceSelection::hoslem.test(MK2_logreg4$y, fitted(MK2_logreg4))

summary(MK2_logreg4)
exp(MK2_logreg4$coefficients)
exp(confint.lm(MK2_logreg4))

# interaction between MK2 and Casp-3

# Exploratory graphs

# Distribution of casp3 transccript data
hist(MK2_CoxPH_MK2andCasp3$casp3.Expression.scaled)

# relationship between MK2 and Casp3)
ggplot(MK2_CoxPH_MK2andCasp3, aes(x=casp3.Expression.scaled, y=clinical.Expression.scaled)) +geom_point()
```

## Cox PH now with right censoring at 1 year. 
```{r}
# COX PH

# data cleaning prior to Cox PH model
MK2_CoxPH_quantile$MK2_Expression_topQ <- as.factor(unlist(MK2_CoxPH_quantile$MK2_Expression_topQ))
MK2_CoxPH_quantile$clinical.gender <- as.factor(MK2_CoxPH_quantile$clinical.gender)

# be careful here. as.numeric alone will return the factor level, not the value!
MK2_CoxPH_quantile$censor <- as.numeric(as.character(MK2_CoxPH_quantile$death_at_1year))

# since the last model involves using casp3, we'll apply the same modif to the df that has mk2 and casp3 exp together
MK2_CoxPH_MK2andCasp3$MK2_Expression_topQ <- as.factor(unlist(MK2_CoxPH_MK2andCasp3$MK2_Expression_topQ))
MK2_CoxPH_MK2andCasp3$clinical.gender <- as.factor(MK2_CoxPH_MK2andCasp3$clinical.gender)
MK2_CoxPH_MK2andCasp3$censor <- as.numeric(as.character(MK2_CoxPH_MK2andCasp3$death_at_1year))


# Cox PH model with MK2 only

MK2_coxph_model_tb_mv_1y <- coxph(Surv(clinical.time, censor) ~ clinical.Expression.scaled:strata(clinical.ajcc_pathologic_stage_combined)+clinical.gender+clinical.smoking, id=clinical.submitter_id, data=MK2_CoxPH_quantile)
# PH testing
cox.zph(MK2_coxph_model_tb_mv_1y)
summary(MK2_coxph_model_tb_mv_1y)
exp(MK2_coxph_model_tb_mv_1y$coefficients)
exp(confint(MK2_coxph_model_tb_mv_1y))


# MK2 model with MK2 and casp3


MK2_coxph_model_tb_mv_1y_casp3 <- coxph(Surv(clinical.time, censor) ~ clinical.Expression.scaled:strata(clinical.ajcc_pathologic_stage_combined)+clinical.gender+clinical.smoking+casp3.Expression.scaled, id=clinical.submitter_id, data=MK2_CoxPH_MK2andCasp3)



MK2_coxph_model1 <- coxph(Surv(clinical.time, censor) ~ clinical.Expression.scaled, id=clinical.submitter_id, data=MK2_CoxPH_MK2andCasp3)
MK2_coxph_model2 <- coxph(Surv(clinical.time, censor) ~ clinical.Expression.scaled+clinical.ajcc_pathologic_stage_combined, id=clinical.submitter_id, data=MK2_CoxPH_MK2andCasp3)
MK2_coxph_model3 <- coxph(Surv(clinical.time, censor) ~ clinical.Expression.scaled+clinical.ajcc_pathologic_stage_combined+clinical.Expression.scaled*clinical.ajcc_pathologic_stage_combined, id=clinical.submitter_id, data=MK2_CoxPH_MK2andCasp3)

summary(MK2_coxph_model_tb_mv_1y_casp3)
exp(MK2_coxph_model_tb_mv_1y_casp3$coefficients)
exp(confint(MK2_coxph_model_tb_mv_1y_casp3))
```

## Graphics-KM curve - here, we're going to stratify MK2 by "hi" vs. "low"
```{r}
library(survminer)
library(survival)
km_early <- survfit(Surv(clinical.time, censor)~MK2_Expression_topQ, data=MK2_CoxPH_quantile[MK2_CoxPH_quantile$clinical.ajcc_pathologic_stage_combined=="Early Stage",], type='kaplan-meier')
ggsurvplot(km_early, xlim=c(0,365), conf.int = TRUE, break.x.by=100, pval=TRUE)

km_late <- survfit(Surv(clinical.time, censor)~MK2_Expression_topQ, data=MK2_CoxPH_quantile[MK2_CoxPH_quantile$clinical.ajcc_pathologic_stage_combined=="Late Stage",], type='kaplan-meier')
ggsurvplot(km_late, xlim=c(0,365), conf.int = TRUE, break.x.by=100, pval=TRUE)

```


# Demographics / Table 1 
```{r}

library(tableone)
table1_df <- left_join(clinical, MK2_CoxPH_quantile[,c(4,14)], by=c("submitter_id"="clinical.submitter_id"))
listVars <- c("age_at_diagnosis", "gender", "smoking", "ajcc_pathologic_stage","ajcc_pathologic_n", "prior_malignancy", "primary_diagnosis", "days_to_last_follow_up", "race")
catVars <- c("gender", "smoking", "ajcc_pathologic_stage", "ajcc_pathologic_n", "prior_malignancy", "primary_diagnosis", "race") 

table1 <- CreateTableOne(listVars, table1_df, catVars, strata=c("MK2_Expression_topQ"))
View(table1)
```