---
title: "Analysis of MK2 transcript levels in TCGA-LUAD and TCGA-LUSC (NSCLC) dataset"
author: "Karthik Suresh"
date: "May 15, 2019"
output:
  html_document:
    theme: united
    toc: yes
  pdf_document:
    toc: yes
---
Here, we are downloading clinical data from TCGA-LUAD dataset using the TCGAbiolinks API. For the MK2 transcript data, I grabbed these data from oncoLnc.org using "MAPKAPK2" as the gene name and TCGA-LUAD as the cohort. 

# Data retrieval and cleaning

## Retrieve data from TCGA dataset, _join it with the MK2 transcript data
```{r set-options, echo=FALSE, cache=FALSE}
options(width = 400)
```

```{r setup, include=TRUE}
knitr::opts_chunk$set(echo = TRUE)
rm(list=ls())
library("SummarizedExperiment")
library("dplyr")
library("DT")
library("TCGAbiolinks")
library(survival)
library(ggplot2)
library(ggpubr)
library(survminer)
library(knitr)
clinical_adeno <- GDCquery_clinic(project = "TCGA-LUAD", type = "clinical")
clinical_squam <- GDCquery_clinic(project="TCGA-LUSC", type = "clinical")
expr <- read.csv("LUAD.MAPKAPK2Exp.csv")
expr_squam<-read.csv("LUSC.MAPAPK2Exp.csv")
clinical_adeno <- left_join(clinical_adeno, expr, c("submitter_id"= "Patient"))
clinical_squam <- left_join(clinical_squam, expr_squam, c("submitter_id"="Patient"))

clinical <- rbind(clinical_adeno, clinical_squam)

library("tidyr")
```

## Clean the data

```{r}
clinical$time <- mapply(function(x,y) if(is.na(x)) return(y) else return(x), clinical$days_to_death, clinical$days_to_last_follow_up) 
clinical$age_at_diagnosis <- clinical$age_at_diagnosis/365
clinical$race <- as.factor(clinical$race)
clinical$smoking <- lapply(clinical$cigarettes_per_day, function(x) if(!is.na(x)) return(1) else return(0))
clinical$tumor_stage <- as.factor(clinical$tumor_stage)
clinical$vital_status <- as.factor(clinical$vital_status)
clinical$status <- lapply(clinical$vital_status, function(x) if(x=="alive") return(0) else return(1))



# combine clinical stage

combineStage <- function(x) {
if(!is.na(x))  {
if(x=="Stage I" | x=="Stage IB" | x=="Stage IA") return("Early Stage")
if(x=="Stage II" | x=="Stage IIA" | x=="Stage IIB") return("Early Stage")
if(x=="Stage IIIA" | x=="Stage IIIB") return("Late Stage")
if(x=="Stage IV") return("Late Stage")
}
  } 

# combined histologic types 

combinePath <- function(x) {
 if(!is.na(x))  {
   
   if(x=="Adenocarcinoma, NOS") return("Adenocarcinoma")
   if(x=="Mucinous adenocarcinoma")  return("Adenocarcinoma")
   if(x=="Adenocarcinoma with mixed subtypes") return("Adenocarcinoma")
   if(x=="Papillary adenocarcinoma") return("Adenocarcinoma")
   if(x=="Squamous cell carcinoma, NOS") return("Squamous Cell Carcinoma")
   else return("Other")
   
 }
  
}

# combine sub-stages into 1,2,3
clinical$ajcc_pathologic_stage_combined <- lapply(clinical$ajcc_pathologic_stage, combineStage)

clinical$ajcc_pathologic_stage_combined <- as.factor(as.character(clinical$ajcc_pathologic_stage_combined))

# combine histology types
clinical$primary_diagnosis_combined <- lapply(clinical$primary_diagnosis, combinePath)
clinical$primary_diagnosis_combined<- as.factor(unlist(clinical$primary_diagnosis_combined))

# get rid of NULL entries
clinical <- clinical[clinical$ajcc_pathologic_stage_combined!="NULL",]
```

# Plot differences in transcript level by tumor histology and cancer stage
```{r}
# plot MK2 differences for adenocarcinoma only
ggplot(clinical[clinical$primary_diagnosis_combined=="Adenocarcinoma", ], aes(x=ajcc_pathologic_stage_combined, y=Expression, ))+geom_boxplot()+geom_point(position=position_jitter())+stat_compare_means()+theme_bw()

ggplot(clinical[clinical$primary_diagnosis_combined=="Squamous Cell Carcinoma", ], aes(x=ajcc_pathologic_stage_combined, y=Expression, ))+geom_boxplot()+geom_point(position=position_jitter())+stat_compare_means()+theme_bw()

ggplot(clinical[clinical$primary_diagnosis_combined!="Other",], aes(fill=ajcc_pathologic_stage_combined, y=Expression, x=primary_diagnosis_combined))+geom_boxplot()+geom_point(position=position_jitterdodge())

# Anova and Tukey HSD
NSCLC_MK2_aov<- aov(data=clinical[clinical$primary_diagnosis_combined!="Other",], Expression~primary_diagnosis_combined + ajcc_pathologic_stage_combined+ajcc_pathologic_stage_combined*primary_diagnosis_combined)

TukeyHSD(NSCLC_MK2_aov)
```


# COX PH looking at survival ~ expression (continuous) + histologic stage (Adeno vs. non-adeno) + clinical stage (early v. late)

```{r}
MK2_CoxPH <-data.frame(clinical$gender, clinical$age_at_index, clinical$submitter_id, clinical$ajcc_pathologic_stage_combined, clinical$primary_diagnosis_combined, clinical$Expression, clinical$time, clinical$vital_status, clinical$age_at_diagnosis)

# We already got rid of NULLs for ajcc_pathologic_stage. However, the factor levels still list 3 levels - early, late and NULL. To re-level, convert to character, re-convert to factor, and now only 2 levels. 
MK2_CoxPH$clinical.ajcc_pathologic_stage_combined<-as.factor(as.character(MK2_CoxPH$clinical.ajcc_pathologic_stage_combined))

MK2_CoxPH <- MK2_CoxPH[!is.na(MK2_CoxPH$clinical.Expression), ]

# create censor variable for CoxPH
MK2_CoxPH$censor <- lapply(MK2_CoxPH$clinical.vital_status, function(x) if(x=="Alive") return(0) else return(1))
MK2_CoxPH$censor <- as.numeric(MK2_CoxPH$censor)

# COX PH model
MK2_coxph_model <- coxph(Surv(clinical.time, censor) ~ clinical.Expression+clinical.ajcc_pathologic_stage_combined+clinical.primary_diagnosis_combined+clinical.age_at_diagnosis, id=clinical.submitter_id, data=MK2_CoxPH)

# examine MK2 expression values 
hist(MK2_CoxPH$clinical.Expression)


# PH assumption testing

PHtesting_MK2contin <- cox.zph(MK2_coxph_model)
PHtesting_MK2contin

# residual examination for individual variables
ggcoxdiagnostics(MK2_coxph_model, type="dfbeta")
ggcoxzph(PHtesting_MK2contin, font.main=c(8,"plain", "black"), font.x=c(8, "plain", "black"), font.y=c(8, "plain", "black"), font.caption=c(8, "plain", "black"), font.tickslab=c(8, "plain", "black"))

summary(MK2_coxph_model)
```

# COX PH looking at survival ~ expression when expression is categorized as upper 33% or lower 33%. 

```{r}
# classify the MK2 expression variable as either top or bottom quantile, based on quantile definitions
bottom_quant <<- 0.33
top_quant <<- 0.66

MK2_CoxPH$MK2_Expression_bottomQ <- lapply(MK2_CoxPH$clinical.Expression, function (x) if(x < quantile(MK2_CoxPH$clinical.Expression, bottom_quant))  return(1) else return(0))

MK2_CoxPH$MK2_Expression_topQ <- lapply(MK2_CoxPH$clinical.Expression, function (x) if(x > quantile(MK2_CoxPH$clinical.Expression, top_quant))  return(1) else return(0))


MK2_CoxPH_tb <- MK2_CoxPH[MK2_CoxPH$MK2_Expression_bottomQ ==1 | MK2_CoxPH$MK2_Expression_topQ==1,]
MK2_CoxPH_tb$MK2_Expression_topQ <- as.numeric(MK2_CoxPH_tb$MK2_Expression_topQ)
MK2_CoxPH_tb$MK2_Expression_bottomQ <- as.numeric(MK2_CoxPH_tb$MK2_Expression_bottomQ)

# re-categorize the variable as MK2_tv where 0 = lower 10th, 1= top 10th
MK2_CoxPH_tb$MK2_tv <- mapply(function(bottomQ, topQ) if(bottomQ==1) return("Low") else return("High"), MK2_CoxPH_tb$MK2_Expression_bottomQ, MK2_CoxPH_tb$MK2_Expression_topQ)
MK2_CoxPH_tb$MK2_tv <- as.factor(MK2_CoxPH_tb$MK2_tv)
MK2_coxph_model_tb <- coxph(Surv(clinical.time, censor) ~ MK2_tv, id=clinical.submitter_id, data=MK2_CoxPH_tb)
PHtesting_MK2categ<- cox.zph(MK2_coxph_model_tb)

# this is not as useful because topQ is a categorical variable. a log-log plot will be more useful

ggcoxzph(PHtesting_MK2categ)

# log-log plot showing lack of proportionality
plot(survfit(Surv(MK2_CoxPH_tb$clinical.time, MK2_CoxPH_tb$censor)~as.numeric(MK2_CoxPH_tb$MK2_tv)), fun="cloglog", xlab="log(t)", ylab="log(-log(Survival))") 
```

# Using a time variable to account for PH assumption violation

We have a categorical variable (MK2 quantile, 0 if bottom 10%, 1 if top 10%) that violates PH. We split the time into chunks. We will arbitrary cut for the step function at t=100,300, 600,1000

```{r}
MK2_CoxPH_tb <- MK2_CoxPH_tb[MK2_CoxPH_tb$clinical.time > 0 & MK2_CoxPH_tb$clinical.primary_diagnosis_combined!="Other",]
MK2_tvary <- survSplit(Surv(clinical.time, censor) ~ ., data=MK2_CoxPH_tb, cut=c(365,600,1000), episode="tgroup",zero=0)

# have a look at the result of split by time chunks at 0-100d (tgroup=1), 100-300d (tgroup=2) and so on...
kable(MK2_tvary[1:20,])

MK2_tvary_model <- coxph(Surv(tstart,clinical.time, censor)~MK2_tv:strata(tgroup):strata(clinical.ajcc_pathologic_stage_combined), data=MK2_tvary[MK2_tvary$clinical.primary_diagnosis_combined!="Squamous Cell Carcinoma",])
MK2_tvary_model

MK2_tvary_model2 <- coxph(Surv(tstart,clinical.time, censor)~MK2_tv:strata(tgroup):strata(clinical.ajcc_pathologic_stage_combined), data=MK2_tvary[MK2_tvary$clinical.primary_diagnosis_combined!="Adenocarcinoma",])
MK2_tvary_model2

MK2_tvary_model3 <- coxph(Surv(tstart,clinical.time, censor)~MK2_tv:strata(tgroup):strata(clinical.ajcc_pathologic_stage_combined):strata(clinical.primary_diagnosis_combined), data=MK2_tvary)
MK2_tvary_model3

exp(confint(MK2_tvary_model))




MK2_tvary_model2 <- coxph(Surv(tstart,clinical.time, censor)~MK2_tv:strata(tgroup):strata(clinical.primary_diagnosis_combined)+clinical.ajcc_pathologic_stage_combined, data=MK2_tvary)


MK2_tvary_model2



```